---
format: 
  revealjs:
    slide-number: c
    theme: [default, custom_theme.scss]
    chalkboard:
      buttons: false
    preview-links: auto
    output-file: "index"
    output-ext: "html"
---

```{r}
#| label: setup
#| include: false
library(here)

here::i_am("presentation.qmd")

source(
  here::here("R", "Config_file.R")
)

source(
  here::here("R", "Make_figures.R")
)

source(
  here::here("R", "Functions", "include_local_figure.R")
)
```

# [Past Human Impact]{.bold .r-fit-text} {.title}

**Ondřej Mottl**

*Department of Botany - 2024-11-12*

## {.inverse}

:::: {.columns}

::: {.column width="30%"}

<br>
<br>
<br>
<br>

```{r}
#| label: earth image
include_local_figure("AI_generated/Earth.jpeg")
```

:::

::: {.column width="70%" .textcenter .center-vertical}

7/8 [global]{.text-color-highlight-primary} Earth system [boundaries]{.text-color-highlight-secondary} are already [exceeded]{.text-color-highlight-tertiary} 

::: {.text-smaller .text-gray}
(Rockström et al. (2023))
:::

:::

::::

::: {.footer}
Image generated by Bing
:::

## {background-image="Materials/Pexels/pexels-justus-menke-3490295-5214145.jpg" background-opacity="0.35"}

<br>
<br>
<br>
<br>

:::: {.textcenter} 

::: { .r-fit-text .bold}
Humans altered [> 75%]{.text-color-highlight-primary} of ice-free land 

::: {.text-smaller .text-gray}
(Ellis, E. C. & Ramankutty, N. (2008))
:::

:::
::::

::: {.footer}
Image by Justus Menke from Pexels
:::

## When?

<br>
<br>

:::: {.columns}

::: {.column width="33%" .fragment}
```{r}
#| label: when-1
include_local_figure("AI_generated/prehistoric_man.jpeg")
```
:::

::: {.column width="33%" .fragment}

```{r}
#| label: when-2
include_local_figure("AI_generated/ricefield.jpeg")
```
:::

::: {.column width="33%" .fragment}

```{r}
#| label: when-3
include_local_figure("AI_generated/industrialization.jpeg")
```

:::

::::

::: {.footer}
Images created by Bing 
:::

## Human vs Climate {background-image="Materials/AI_generated/human_vs_climate.jpeg" background-opacity="0.50"}

<br>
<br>

:::{.blockquote .fragment .textcenter}
At a larger scale, has [humans]{.text-color-highlight-human .bold} or [climate]{.text-color-highlight-climate .bold}  been the main driver of vegetation change during the early-to-mid Holocene?
:::

::: {.footer}
Image created by Bing 
:::

# Ondřej Mottl

::: {.textcenter}

Head of 🧑‍💻[Laboratory of Quantitative Ecology](https://ondrejmottl.github.io/lab/about_the_lab.html)

:::

:::: {.columns}

::: {.column width="60%"}
<br>

* 📧 Ondrej.mottl(at)gmail.com
* 🐦 [@OndrejMottl](https://twitter.com/OndrejMottl)
* 🐱 [Github OndrejMottl](https://github.com/OndrejMottl)
* 🆔 [ORCID 0000-0002-9796-5081](https://orcid.org/0000-0002-9796-5081)
* 🌐 [bit.ly/ondrej_mottl](https://bit.ly/ondrej_mottl)
:::

::: {.column width="40%"}
```{r}
#| label: head photo
include_local_figure("head_photo.png")
```
:::

::::

# Past Vegetation

```{r}
#| label: pollen-map-full
include_local_figure("R_generated/pollen_map_full.png")
```

::: {.footer}
over 1260 Fossil pollen records from the Neotoma database; over 75,000 individual assemblages
:::

## Europe

```{r}
#| label: pollen-map-europe
include_local_figure("R_generated/pollen_map_europe.png")
```

::: {.footer}
Koppen-Geiger climate zones
:::

## Temperate climate zone

```{r}
#| label: pollen-map-temperate
include_local_figure("R_generated/pollen_map_europe_temperate.png")
```

## Example record

```{r}
#| label: pollen-map-example-record
include_local_figure("R_generated/pollen_map_example_record.png")
```

## Ecosystem Properties

:::: {.columns}

::: {.column width="20%"}

<br>
<br>
<br>
<br>

```{r}
#| label: icon-ecosystem-1
include_local_figure("Preprint/Icons/icon_ecosystem.png")
```
:::

::: {.column width="80%" .fragment .textcenter} 
```{r}
#| label: pollen-paps-1
include_local_figure("R_generated/pollen_paps_all.png")
```

Reflecting the multidimensional aspects of biodiversity
:::

::::

## Climate

:::: {.columns}

::: {.column width="20%"}

<br>
<br>

```{r}
#| label: climate-icon-1
include_local_figure("Preprint/Icons/icon_climate.png")
```
:::

::: {.column width="80%" .fragment .textcenter}
```{r}
#| label: climate-1
include_local_figure("R_generated/climate_line.png")
```

Capturing variation in both temperature and precipitation
:::

::::

## Human presence/impact {auto-animate="true"}

:::: {.columns}

::: {.column width="20%"}

<br>
<br>

```{r}
#| label: human-icon-1
include_local_figure("Preprint/Icons/icon_human.png")
```
:::

::: {.column width="10%"}
:::

::: {.column width="60%" .fragment}

```{r}
#| label: human-map-1
include_local_figure("R_generated/map_example_record_bare.png")
```

:::

::::

## Human presence/impact {auto-animate="true"}

:::: {.columns}

::: {.column width="20%"}

<br>
<br>

```{r}
#| label: human-icon-2
include_local_figure("Preprint/Icons/icon_human.png")
```


```{r}
#| label: human-artefacts-icon-1
include_local_figure("Preprint/Icons/icon_artefatcs.png")
```
:::

::: {.column width="10%"}
:::

::: {.column width="60%" .textcenter}

```{r}
#| label: human-map-2
include_local_figure("R_generated/map_example_record_rc.png")
```

over 150,000 radiocarbon dates derived from archaeological data

:::

::::


## Human presence/impact

:::: {.columns}

::: {.column width="20%"}

<br>
<br>

```{r}
#| label: human-icon-3
include_local_figure("Preprint/Icons/icon_human.png")
```
:::

::: {.column width="80%"}
```{r}
#| label: human-density-1
include_local_figure("R_generated/human_density.png")
```

:::

::::

::: {.footer}
Summed Probability Distribution (SPD)
:::

## Example record - predictor importance

```{r}
#| label: combined-line-1
include_local_figure("R_generated/example_combined.png")
```

## Example record - predictor importance

```{r}
#| label: predictor-importance-1
include_local_figure("R_generated/example_with_importance.png")
```

::: {.footer}
Hiearchical variation partitioning (hVarPart)
:::

## Summary - Europe {auto-animate="true"}

::: {.columns}

::: {.column width="50%"}

```{r}
#| label: summary-map-1
include_local_figure("R_generated/map_example_record_bare.png")
```

:::

::: {.column width="10%"}
:::

::: {.column width="30%"}

```{r}
#| label: summary-values-1
include_local_figure("R_generated/summary_example_record.png")
```

:::

::::

## Summary - Europe {auto-animate="true"}

::: {.columns}

::: {.column width="50%"}

```{r}
#| label: summary-map-2
include_local_figure("R_generated/pollen_map_europe_temperate.png")
```

:::

::: {.column width="10%"}
:::

::: {.column width="30%"}

```{r}
#| label: summary-values-2
include_local_figure("R_generated/summary_eu_temperate.png")
```

:::

::::

## Summary - Europe {auto-animate="true"}

::: {.columns}

::: {.column width="50%"}

```{r}
#| label: summary-map-3
include_local_figure("R_generated/pollen_map_europe_temperate.png")
```

:::

::: {.column width="10%"}
:::

::: {.column width="30%"}

```{r}
#| label: summary-values-3
include_local_figure("R_generated/summary_eu_temperate_quantile.png")
```

:::

::::

## Summary - Europe {auto-animate="true"}

::: {.columns}

::: {.column width="50%"}

```{r}
#| label: summary-map-4
include_local_figure("R_generated/pollen_map_europe.png")
```

:::

::: {.column width="50%"}

```{r}
#| label: summary-values-4
include_local_figure("R_generated/summary_eu_quantile.png")
```

:::

::::

## Summary - Europe {auto-animate="true"}

::: {.columns}

::: {.column width="50%"}

```{r}
#| label: summary-map-5
include_local_figure("R_generated/pollen_map_europe.png")
```

:::

::: {.column width="50%"}

```{r}
#| label: summary-values-5
include_local_figure("R_generated/summary_eu_with_density.png")
```

:::

::::

## Summary - Europe {auto-animate="true"}

::: {.columns}

::: {.column width="50%"}

```{r}
#| label: summary-values-6
include_local_figure("R_generated/summary_eu_with_density.png")
```


:::

::: {.column width="50%"}

```{r}
#| label: summary-values-7
include_local_figure("R_generated/summary_eu_with_density_climate.png")
```

:::

::::

## Summary - World

```{r}
#| label: summary-world
include_local_figure("Preprint/Figure2_h1_spatial_manual_edit.png")
```

# Main findings {.inverse auto-animate="true"}

<br>

::: {.textcenter}
We have summaried over [75,000 vegetation assemblages]{.text-color-highlight-primary} in over 1260 fossil pollen records. Decribe each ecosystem by [9 ecosystem properties]{.text-color-highlight-secondary}. We have also included over [150,000 radiocarbon dates]{.text-color-highlight-human} derived from archaeological data and [4 past climate variables]{.text-color-highlight-climate}.
:::

## Main findings {.inverse auto-animate="true"}

::: {.textcenter .text-color-gray}
We have summaried over 75,000 vegetation assemblages in over 1260 fossil pollen records. Decribe each ecosystem by 9 ecosystem properties. We have also included over 150,000 radiocarbon dates derived from archaeological data and 4 past climate variables.
:::

<br>

::: {.textcenter}
This is the first study to [quantitatively]{.text-color-highlight-primary} estimate the relative importance of [human]{.text-color-highlight-human} and [climate]{.text-color-highlight-climate} effects on vegetation properties globaly and during the early-to-mid [Holocene]{.text-color-highlight-secondary}.
:::


## Main findings {.inverse auto-animate="true"}

::: {.textcenter .text-color-gray}
We have summaried over 75,000 vegetation assemblages in over 1260 fossil pollen records. Decribe each ecosystem by 9 ecosystem properties. We have also included over 150,000 radiocarbon dates derived from archaeological data and 4 past climate variables.
:::

::: {.textcenter .text-color-gray} 
This is the first study to quantitatively estimate the relative importance of human and climate effects on vegetation properties globaly and during the early-to-mid Holocene.
:::

<br>

::: {.textcenter}
[Localy]{.text-color-highlight-primary}, [humans]{.text-color-highlight-human .bold} had [severe impact]{.text-color-highlight-tertiary} on vegetation change during the early-to-mid [Holocene]{.text-color-highlight-secondary}.
:::

## Main findings {.inverse auto-animate="true"}

::: {.textcenter .text-color-gray}
We have summaried over 75,000 vegetation assemblages in over 1260 fossil pollen records. Decribe each ecosystem by 9 ecosystem properties. We have also included over 150,000 radiocarbon dates derived from archaeological data and 4 past climate variables.
:::

::: {.textcenter .text-color-gray} 
This is the first study to quantitatively estimate the relative importance of human and climate effects on vegetation properties globaly and during the early-to-mid Holocene.
:::

::: {.textcenter .text-color-gray}
Localy, humans had severe impact on vegetation change during the early-to-mid Holocene.
:::

<br>

::: {.textcenter}
[At a larger scale]{.text-color-highlight-primary}, [climate]{.text-color-highlight-climate .bold} was the [main driver]{.text-color-highlight-tertiary} of vegetation change during the early-to-mid [Holocene]{.text-color-highlight-secondary}.
:::

## Preprint available

:::: {.columns}

::: {.column width="70%"}

```{r}
#| label: preprint
include_local_figure("Preprint/preprint_header.png")
```

:::

::: {.column width="30%"}

```{r}
#| label: preprint-qr
include_local_figure("QRs/qr_preprint.png")
```

:::

::::

::: {.textcenter .fragment}
[Vivian A. Felde & [Ondřej Mottl]{.text-color-highlight-tertiary} & Suzette G. A. Flantua]{.underline}, Kuber P. Bhatta, Hilary H. Birks, John-Arvid Grytnes, Henry Hooghiemstra, Sandra Nogué, Triin Reitalu, Alistair W. R. Seddon, Manuel J. Steinbauer, Cajo J. F. ter Braak, Thomas Giesecke, Simon J. Goring, Simon G. Haberle, Annika V. Herbert, Petr Šmilauer, Willy Tinner, John W. Williams, Darcy Bird1, Jacob Freeman, Eugenia Gayo, Julie A. Hoggarth, Claudio Latorre, Lux Miranda, Erick Robinson, Xianyong Cao, Fahu Chen, H. John B. Birks
:::

## [This presentation]{.r-fit-text} {.title}

::: {.r-fit-text}
😸 Code on [OndrejMottl GitHub](https://github.com/OndrejMottl/)

[PastHumanImpact-FlashPresentation](https://github.com/OndrejMottl/PastHumanImpact-FlashPresentation-20241112)

[![](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
:::

:::: {.columns}

::: {.column width="33%"}
:::

::: {.column width="33%"}

```{r}
#| label: qr_github
include_local_figure("QRs/qr_github.png")
```
:::

::: {.column width="33%"}
:::

::::

#

::: {.textcenter .bold}
[Climate]{.text-color-highlight-climate} outweighs [human]{.text-color-highlight-human} effects on [vegetation properties]{.text-color-highlight-primary} during the early-to-mid [Holocene]{.text-color-highlight-secondary}
:::

:::: {.columns}

::: {.column width="60%"}

:::: {.columns}

::: {.column width="40%"}

```{r}
#| label: head logo
include_local_figure("Ondrej_Mottl-transparent-shadow.png")
```
:::

::: {.column width="40%"}

```{r}
#| label: qr_ondrejmottl
include_local_figure("QRs/qr_ondrejmottl.png")
```
:::

::::

* 📧 Ondrej.mottl(at)gmail.com
* 🐦 [@OndrejMottl](https://twitter.com/OndrejMottl)
* 🌐 [bit.ly/ondrej_mottl](https://bit.ly/ondrej_mottl)
* 🧑‍💻[Laboratory of Quantitative Ecology](https://ondrejmottl.github.io/lab/about_the_lab.html)

[Department of Botany - 2024-11-12]{.text-color-gray}

:::

::: {.column width="40%"}

```{r}
#| label: head photo 2
include_local_figure("head_photo.png")
```
:::

::::

# [Appendix]{.r-fit-text} {.title}

## Conceptual workflow figure

```{r}
#| label: workflow
include_local_figure("Preprint/Figure1.png")
```

## More examples

```{r}
#| label: more-examples
include_local_figure("Preprint/Extended_data_figure_3_manual_edits.png")
```

## Change in time

```{r}
#| label: change-in-time
include_local_figure("Preprint/Figure3_h1_temporal_manual_edit.png")
```

##  Internal relationships

```{r}
#| label: internal-relationships
include_local_figure("Preprint/Figure4_h2_manual_edit.png")
```

## Pollen data overview

```{r}
#| label: pollen-data-overview
include_local_figure("Preprint/Extended_data_figure_1.png")
```

## Human impact overview

```{r}
#| label: human-impact-overview
include_local_figure("Preprint/Extended_data_figure_2.png")
```

## General patterns - SPD

```{r}
#| label: general-patterns-spd
include_local_figure("Preprint/Extended_data_figure_4_spd.png")
```

## General patterns - Events

```{r}
#| label: general-patterns-events
include_local_figure("Preprint/Extended_data_figure_9.png")
```

## General patterns - MAT

```{r}
#| label: general-patterns-mat
include_local_figure("Preprint/Extended_data_figure_5_temp_annual.png")
```

## General patterns - MTCO

```{r}
#| label: general-patterns-mtco
include_local_figure("Preprint/Extended_data_figure_6_temp_cold.png")
```

## General patterns - MAPS

```{r}
#| label: general-patterns-maps
include_local_figure("Preprint/Extended_data_figure_7_prec_summer.png")
```

## General patterns - MAPW

```{r}
#| label: general-patterns-mapw
include_local_figure("Preprint/Extended_data_figure_8_prec_winter.png")
```

