---
format: 
  revealjs:
    slide-number: true
    theme: [default, custom_theme.scss]
    chalkboard:
      buttons: false
    preview-links: auto
    output-file: "index"
    output-ext: "html"
---

```{r}
#| label: setup
#| include: false
library(here)

here::i_am("presentation.qmd")

source(
  here::here("R", "Config_file.R")
)

source(
  here::here("R", "Make_figures.R")
)

source(
  here::here("R", "Functions", "include_local_figure.R")
)
```

# [Past Human Impact]{.bold .r-fit-text} {.title}

**Ondřej Mottl**

*Department of Botany - 2024-11-12*

## {.inverse}

:::: {.columns}

::: {.column width="30%"}

<br>
<br>
<br>
<br>

```{r}
#| label: earth image
include_local_figure("Earth.jpeg")
```

:::

::: {.column width="70%" .textcenter .center-vertical}

7/8 [global]{.text-color-highlight-primary} Earth system [boundaries]{.text-color-highlight-secondary} are already [exceeded]{.text-color-highlight-tertiary} 

::: {.text-smaller .text-gray}
(Rockström et al. (2023))
:::

:::

::::

::: {.footer}
Image generated by Bing
:::

## {background-image="Materials/pexels-justus-menke-3490295-5214145.jpg" background-opacity="0.35"}

<br>
<br>
<br>
<br>

:::: {.textcenter} 

::: { .r-fit-text .bold}
Humans altered [> 75%]{.text-color-highlight-primary} of ice-free land 

::: {.text-smaller .text-gray}
(Ellis, E. C. & Ramankutty, N. (2008))
:::

:::
::::

::: {.footer}
Image by Justus Menke from Pexels
:::

## When?

<br>
<br>

:::: {.columns}

::: {.column width="33%" .fragment}
```{r}
#| label: when-1
include_local_figure("prehistoric_man.jpeg")
```
:::

::: {.column width="33%" .fragment}

```{r}
#| label: when-2
include_local_figure("ricefield.jpeg")
```
:::

::: {.column width="33%" .fragment}

```{r}
#| label: when-3
include_local_figure("industrialization.jpeg")
```

:::

::::

::: {.footer}
Images created by Bing 
:::

## Human vs Climate {background-image="Materials/human_vs_climate.jpeg" background-opacity="0.50"}

<br>
<br>

:::{.blockquote .fragment .textcenter}
At a larger scale, has [humans]{.text-color-highlight-human .bold} or [climate]{.text-color-highlight-climate .bold}  been the main driver of vegetation change during a Holocene?
:::

::: {.footer}
Image created by Bing 
:::

# Ondřej Mottl

::: {.textcenter}

Head of 🧑‍💻[Laboratory of Quantitative Ecology](https://ondrejmottl.github.io/lab/about_the_lab.html)

:::

:::: {.columns}

::: {.column width="60%"}
<br>

* 📧 Ondrej.mottl(at)gmail.com
* 🐦 [@OndrejMottl](https://twitter.com/OndrejMottl)
* 🐱 [Github OndrejMottl](https://github.com/OndrejMottl)
* 🆔 [ORCID 0000-0002-9796-5081](https://orcid.org/0000-0002-9796-5081)
* 🌐 [bit.ly/ondrej_mottl](https://bit.ly/ondrej_mottl)
:::

::: {.column width="40%"}
```{r}
#| label: head photo
include_local_figure("head_photo.png")
```
:::

::::

## Past Vegetation

```{r}
#| label: pollen-map-full
include_local_figure("R-generated/pollen_map_full.png")
```

::: {.footer}
over 1260 Fossil pollen records from the Neotoma database; over 75,000 individual assemblages
:::

## Europe

```{r}
#| label: pollen-map-europe
include_local_figure("R-generated/pollen_map_europe.png")
```

## Temperate climate zone

```{r}
#| label: pollen-map-temperate
include_local_figure("R-generated/pollen_map_europe_temperate.png")
```

## Example record

```{r}
#| label: pollen-map-example-record
include_local_figure("R-generated/pollen_map_example_record.png")
```

## Ecosystem Properties

:::: {.columns}

::: {.column width="20%"}

<br>
<br>
<br>
<br>

```{r}
#| label: icon-ecosystem-1
include_local_figure("icon_ecosystem.png")
```
:::

::: {.column width="80%" .fragment}
```{r}
#| label: pollen-paps-1
include_local_figure("R-generated/pollen_paps_all.png")
```
reflecting the multidimensional aspects of biodiversity
:::

::::

## Climate

:::: {.columns}

::: {.column width="20%"}

<br>
<br>

```{r}
#| label: climate-icon-1
include_local_figure("icon_climate.png")
```
:::

::: {.column width="80%" .fragment}
```{r}
#| label: climate-1
include_local_figure("R-generated/climate_line.png")
```

capturing variation in both temperature and precipitation
:::

::::

## Human presence/impact {auto-animate="true"}

:::: {.columns}

::: {.column width="20%"}

<br>
<br>

```{r}
#| label: human-icon-1
include_local_figure("icon_human.png")
```
:::

::: {.column width="80%" .fragment}

```{r}
#| label: human-map-1
include_local_figure("R-generated/map_example_record_bare.png")
```

:::

::::

## Human presence/impact {auto-animate="true"}

:::: {.columns}

::: {.column width="20%"}

<br>
<br>

```{r}
#| label: human-icon-2
include_local_figure("icon_human.png")
```
:::

::: {.column width="80%"}

```{r}
#| label: human-map-2
include_local_figure("R-generated/map_example_record_rc.png")
```

:::

::::

::: {.footer}
over 150,000 radiocarbon dates derived from archaeological data
:::

## Human presence/impact

:::: {.columns}

::: {.column width="20%"}

<br>
<br>

```{r}
#| label: human-icon-3
include_local_figure("icon_human.png")
```
:::

::: {.column width="80%"}
```{r}
#| label: human-density-1
include_local_figure("R-generated/human_density.png")
```

:::

::::

::: {.footer}
Summed Probability Distribution (SPD)
:::

## Example record - predictor importance

```{r}
#| label: combined-line-1
include_local_figure("R-generated/example_combined.png")
```

## Example record - predictor importance

```{r}
#| label: predictor-importance-1
include_local_figure("R-generated/example_with_importance.png")
```

## [This presentation]{.bold .r-fit-text} {.title}

::: {.r-fit-text}
[😸 Code on GitHub](https://github.com/OndrejMottl/PastHumanImpact-FlashPresentation-20241112)

[🖥️ Slides](https://ondrejmottl.github.io/PastHumanImpact-FlashPresentation-20241112/)

[![](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
:::





